Difference Scores: Simple Regression

### 2024-05-01 examples/1a_Examples_DifferenceScores_PairedTtests.qmd
update example 1b

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

>>>>

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = show,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

### 2024-05-01 examples/1b_Examples_DifferenceScores_SimpleRegression.qmd
update example 1b

# Simple Regression {.unnumbered}

## Overview

<p>

This example assesses whether growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1) differs significantly based on handedness, using height ("anthroheightcalc") as a representative metric of growth. The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 height measurements for each participant; 2) a simple regression analysis is used to test whether sex (boy, girl) predicts the average difference value in participants height from baseline to the 1-Year follow-up. Finally, a visual inspection is further conducted via a violin plot to graphically represent the relationship between difference scores and sex. The ensuing analysis and interpretations are detailed in the subsequent sections.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# rm(list = ls())

# Create a list of required packages
packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
library(rstatix)      # Pipe-friendly framework for performing common stats
library(gtsummary)    # Publication-ready tables
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page

```
:::

```{r config}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

### Config Options
#::: blue
#**This code configures knitr code chunk options**

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
                 anthroweightcalc ~ "Weight"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```

:::
:::

## Results

::: panel-tabset
### Build Model: Step 1 {.tabset .tabset-fade .tabset-pills}

::: blue
The code snippet below tells R to compute a difference score by subtracting each participant’s height at T1 from their height at T0. Relevant summary statistics are also provided.

**STEP 1: Compute Difference Score**
```{r build-model-regression-1}
#| message: FALSE
#| warning: FALSE
#| echo: TRUE
#| class.source: 'fold-hide'

# Define the function to compute difference scores for a given variable and provide a summary
# Function to compute difference scores for a given variable and provide a summary
compute_difference_and_summary <- function(df, variable_name) {
    # Define the event names of interest
    baseline_event <- "baseline_year_1_arm_1"
    followup_event <- "1_year_follow_up_y_arm_1"

    # Compute the difference between Baseline and Year 1 data for the given variable
    diff_data <- df %>%
        filter(eventname %in% c(baseline_event, followup_event)) %>% # Filter for specific event names
        select(src_subject_id, eventname, all_of(variable_name)) %>% # Select required columns
        spread(eventname, variable_name) %>% # Convert data from long to wide format
        mutate(diff = get(followup_event) - get(baseline_event)) %>% # Compute difference between the two time points
        drop_na(diff) # Exclude rows with NA in the computed difference

    # Summarize the computed difference scores
    diff_summary <- summary(diff_data$diff)

    # Return the difference data and its summary
    list(data = diff_data, summary = diff_summary)
}

# List of variables for which difference scores are to be computed
variables_of_interest <- c("anthroheightcalc")

# Compute and store difference scores and summaries for each variable in a list
difference_and_summary_list <- lapply(variables_of_interest, function(var) {
    compute_difference_and_summary(df, var)
})

# Extract the difference data for the 'anthroheightcalc' variable
height_diff_data <- difference_and_summary_list[[1]]$data

# Merge the 'diff' column back to the main df using 'src_subject_id' as the key
df <- left_join(df, height_diff_data %>% select(src_subject_id, diff), by = "src_subject_id")

@@

**Descriptive Statistics for the Difference Score**
```{r model-checking}
# Compute statistical summaries for the difference score variable
lapply(difference_and_summary_list, function(item) {
    print(item$summary)
})
```

This summary of the difference score variable indicates that the differences range from a decrease of `r min(height_diff_data$diff, na.rm = TRUE)` units to an increase of `r max(height_diff_data$diff, na.rm = TRUE)` units. The median difference is `r median(height_diff_data$diff, na.rm = TRUE)` units, and the average difference is approximately `r mean(height_diff_data$diff, digits = 2, na.rm = TRUE)` units. There are `r sum(is.na(height_diff_data$diff))` missing values in this difference score dataset.

**Descriptive Statistics for the Height Variable**
```{r}
## Summary statistics
# Compute summary statistics for Height by eventname
summary <- df %>%
    group_by(eventname) %>%
    get_summary_stats(anthroheightcalc, type = "mean_sd")

data.frame(summary)
```

The summary statistics provide insights into the Height variable across different `eventname` categories. At baseline, the average height is approximately `r summary$mean[summary$eventname == "baseline_year_1_arm_1"]` units with a standard deviation of `r summary$sd[summary$eventname == "baseline_year_1_arm_1"]`. Over the years, there’s a noticeable increase in average height: by the first year, it’s about `r summary$mean[summary$eventname == "1_year_follow_up_y_arm_1"]` units, and it continues to rise, reaching approximately `r summary$mean[summary$eventname == "4_year_follow_up_y_arm_1"]` units by the fourth year. The standard deviation remains relatively consistent over the years, suggesting similar variability in height measurements across these time points.

:::

### Build Model: Step 2 {.tabset .tabset-fade .tabset-pills}

::: blue
A simple regression analyses is conducted to examine whether a grouping variable (sex) significantly predicts the difference score value (indicating significant group differences in the average difference score).

**STEP 2: Conduct regression on Difference Score**
```{r build-model-regression-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
merged_data <- height_diff_data %>%
  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")

# Ensure 'sex' is a factor
merged_data$sex <- as.factor(merged_data$sex)

# Run the regression
model <- lm(diff ~ sex, data = merged_data)

# Get the summary of the regression model
model_summary <- summary(model)

model_summary

@@

This regression analysis evaluates whether sex (boy, girl) predicts the average difference in participants' height from baseline to the 1-Year follow-up. The output from our model provides:

i. an F-statistic of `r model_summary$fstatistic[1]`; 
ii. degrees of freedom of `r model_summary$df[2]`; 
iii. a parameter estimate of `r model_summary$coefficients["sexMale or other", "Estimate"]`; 
iv. standard error of `r model_summary$coefficients["sexMale or other", "Std. Error"]`; 
v. p-value of `r model_summary$coefficients["sexMale or other", "Pr(>|t|)"]`.

Compared to girls (the reference group), boys have an average height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units. This effect was statistically significant with a p-value of `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`. Overall, the model explained a very small portion of the variance in height difference, with an adjusted R-squared value of `r round(model_summary$adj.r.squared, 6)`, and the overall model significance was  statistically significant with a p-value of `r pf(model_summary$fstatistic[1], model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)`.

:::

### Model Plots

::: blue
```{r violin-plots}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

# Visualize the difference scores across different levels of sex

# Create a violin plot to show the distribution of difference scores by sex
# Jittered points are added to provide a more granular view of individual observations
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# Merge the 'sex' variable from the original dataframe 'df' to 'height_diff_data'
merged_data <- height_diff_data %>%
  left_join(df %>% select(src_subject_id, sex), by = "src_subject_id")

# Plotting using ggplot2
ggplot(merged_data, aes(x = sex, y = diff, fill = sex)) +
  geom_violin() +
  geom_jitter(position = position_jitter(width = 0.2), size = 1, alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") + 
  labs(
    title = "Difference Scores by Sex",
    x = "Sex",
    y = "Difference Score"
  ) +
  theme_minimal() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

@@

The violin plot visualizes the distribution of difference scores in children's heights by their sex: Right, Left, and Mixed. Each violin shape provides insight into the density of the data at different height differences, with wider sections representing higher densities of data points. Superimposed jittered points offer a granular view of individual observations. From the plot, it appears that the distributions of height differences across the three handedness categories are somewhat similar, though there might be subtle variations in median and spread.

:::
:::

## Wrapping Up

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue
Differences in children's heights between baseline and a subsequent 1-year follow-up, we sought to understand the potential influence of sex on this difference. Descriptive statistics revealed that height differences ranged from a decrease of `r min(merged_data$diff, na.rm = TRUE)` units to an increase of `r max(merged_data$diff, na.rm = TRUE)` units, with a median difference of `r median(merged_data$diff, na.rm = TRUE)` units and an average difference of approximately `r mean(merged_data$diff, na.rm = TRUE)` units. Subsequent regression analysis indicated that, when compared to boys, girls had an average increase in height difference of approximately `r round(model_summary$coefficients["sexMale or other", "Estimate"], 3)` units, though this effect was marginally significant (p = `r round(model_summary$coefficients["sexMale or other", "Pr(>|t|)"], 4)`). Complementary to these findings, violin plots visually underscored the subtle variations in height differences across sex categories, suggesting broadly similar distributions but with nuanced variations in central tendency and spread. In conclusion, while sex exhibited a potential influence on height difference over the year, the observed effect was notably minor.

:::
:::

>>>>

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = show,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, ehi_y_ss_scoreb, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    group_by(src_subject_id) %>%
    fill(ehi_y_ss_scoreb, .direction = "downup") %>%
    ungroup() %>%
    filter(ehi_y_ss_scoreb %in% c(1, 2)) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1")),
        ehi_y_ss_scoreb = factor(ehi_y_ss_scoreb, levels = c(1, 2), labels = c("Right", "Left")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        handedness = ehi_y_ss_scoreb,
        height = anthroheightcalc
    )

@@

## Overview
Difference scores can be used to assess change in a variable across two timepoints. The value of the variable at the first timepoint is substracted from the value at the second timepoint. A simple regression can then be applied to determine whether a variable of interest predicts the computed difference score.

This example will examine height values obtained across 2 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to determine whether a variable of interest (handedness) predicts the difference in height values between the two timepoints. A difference score will be calculated between participant's height values collected at baseline and again at the 1-year follow-up. Next, a simple regression analysis will be conducted to test whether handedness (left, right) predicts the average difference value in participants' height from baseline to the 1-Year follow-up. Finally, a visual inspection is further conducted via a violin plot to graphically represent the relationship between the difference score and handedness.

## Preliminary Setup
### Load R libraries
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format

@@

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, handedness, height) %>%
    mutate(event = factor(event)) %>%
    mutate(handedness = factor(handedness)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(handedness ~ "Handedness", height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

@@

## Results
### Compute Difference Score
This code computes a difference score by subtracting participant's height at baseline from height at the 1-year follow-up. Next, a simple regression analysis is conducted to determine whether handedness predicts the difference in height values between the two timepoints.

```{r}

# Reshape data from long to wide format
df_wide <- df_long %>%
  pivot_wider(
    id_cols = c(id, handedness),
    names_from = event,
    values_from = height,
    names_prefix = "Height_"
  )

# Compute the height difference score
df_wide <- df_wide %>%
    mutate(
        height_diff = Height_Year_1 - Height_Baseline
    )

# Calculate summary statistics for the difference score
summary(df_wide$height_diff)

# Compute summary statistics for Height by eventname
summary <- df_long %>%
    group_by(event) %>%
    get_summary_stats(height, type = "mean_sd")

data.frame(summary)

```
The summary statistics provide insights into participant's height across two annual assessments. At baseline, the average height is 55.24 inches (sd = 3.33). Over the year, there is an increase of 2.35 inches in average height to 57.59 inches at the time of the 1-year follow-up.

### Conduct regression on Difference Score
A simple regression analyses is conducted to examine whether a grouping variable (handedness) predicts the difference score value.
```{r}

# Run the regression
model <- lm(height_diff ~ handedness, data = df_wide)

# Get the summary of the regression model
summary(model)

The results from the simple regression analysis suggest handedness does not significantly predict change in height. The intercept, estimated at 2.35 with a standard error of .02, indicates the average height difference for right-handed individuals is approximately 2.35 units, with this effect differing by only .03 inches (p=.69) for left-handed individuals

### Model Plots
```{r}

# Visualize the difference scores across different levels of handedness

# Create a violin plot to show the distribution of difference scores by sex
# Jittered points are added to provide a more granular view of individual observations
# Load necessary libraries
library(RColorBrewer)

# Plotting using ggplot2
ggplot(df_long, aes(x = handedness, y = height, fill = handedness)) +
    geom_violin() +
    geom_jitter(position = position_jitter(width = 0.2), size = 1, alpha = 0.5) +
    scale_fill_brewer(palette = "Set2") +
    labs(
        title = "Difference Scores by Handedness",
        x = "Handedness",
        y = "Height Difference Score"
    ) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    )

The violin plot visualizes the distribution of difference scores in children's heights by their handedness (Right, Left). Each violin shape provides insight into the density of the data at different height differences, with wider sections representing higher densities of data points. Superimposed jittered points offer a granular view of individual observations. From the plot, it appears that the distributions of height differences across the two handedness categories are markedly similar.

## Wrapping Up
This analysis examined difference scores of participant's height across a one-year period and used a simple regression analysis to explore the impact of handedness on the year-to-year change in children's height. Findings showed an increase in height values from baseline (55.23 inches) to the year 1 follow-up (57.57 inches). The regression model's output shows that handedness, specifically being left-handed as compared to right-handed, is not associated with any change in height (r=.03, se=.08, p=.69). Further, a violin plot visualizing the distribution of difference scores in children's heights by their handedness showed no substantive effect.

This difference score approach used in this example quantifies the change in height values over time and uses a simple regression to test whether participant handedness predicts this value. More generally, this approach is often implemented when data used to evalaute change is only available at two timepoints, as more flexible approaches are typically implemented when data from additional (>2) time points is available.
